---
title: "hw1"
author: "Lai Wei"
format:
  html:
    embed-resources: true
editor: visual
---

The link to the GitHub repository of this assignment is: https://github.com/Lwei493/Homework1

## Problem 1 - Exploratory Analysis of Soybean Cultivars

# a.

Import the data

```{r}
#| echo: true
soybean <- read.table("data/forty+soybean+cultivars+from+subsequent+harvests/data.csv", sep = ",", header = TRUE)
```

# b.

Using the information provided in the dataset documentation: Verify that there are 320 observations in the dataset; 40 distinct cultivars; all the observations were collected from two seasons. Produce a table showing the number of observations from each season.

```{r}
#| echo: true
total_obs <- nrow(soybean)
num_cultivars <- length(unique(soybean$Cultivar))
num_seasons <- length(unique(soybean$Season))
season_table <- table(soybean$Season)
cat("The total observations are", nrow(soybean),
    "\nNumber of distinct cultivars is", length(unique(soybean$Cultivar)),
    "\nNumber of seasons is", length(unique(soybean$Season)), "\n")
season_table
```

# c.

1.  What is the correlation between plant height (PH) and grain yield (GY)?

```{r}
#| echo: true
corr <- cor(soybean$PH, soybean$GY, use = "complete.obs")
cat("Correlation between PH and GY is", round(corr, 4), "\n\n")
```

2.  Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?

```{r}
#| echo: true
cor_season1 <- cor(soybean$PH[soybean$Season == 1], soybean$GY[soybean$Season == 1], use = "complete.obs")
cor_season2 <- cor(soybean$PH[soybean$Season == 2], soybean$GY[soybean$Season == 2], use = "complete.obs")
cat("Correlation between PH and GY in Season 1 is", round(cor_season1, 4), "\n")
cat("Correlation between PH and GY in Season 2 is", round(cor_season2, 4), "\n")
if (cor_season1 > cor_season2) {cat("Season 1 has the higher correlation.\n\n")
  }else if (cor_season2 > cor_season1) {cat("Season 2 has the higher correlation.\n\n")
  } else {cat("Both seasons have equal correlation.\n\n")}
```

3.  What is the grain yield of the soybean observation with the greatest plant height?

```{r}
#| echo: true
max_PH <- max(soybean$PH, na.rm = TRUE)
GY_with_maxPH <- soybean$GY[which.max(soybean$PH)]
cat("Grain yield (GY) at the observation with the greatest plant height is", round(GY_with_maxPH, 2), "\n\n")
```

4.  What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

```{r}
#| echo: true
average_MHG <- mean(soybean$MHG, na.rm = TRUE)
above_average <- sum(soybean$MHG > average_MHG, na.rm = TRUE)
percent_above <- above_average / total_obs * 100
cat("Overall average MHG is", round(average_MHG, 2),
    "\nObservations with MHG > average:", above_average, "out of", total_obs,
    "\nPercentage =", round(percent_above, 2), "%")
```

5.  Which cultivar has the highest average grain yield?

```{r}
#| echo: true
average_GY_by_cultivar <- aggregate(GY ~ Cultivar, data = soybean, FUN = mean, na.rm = TRUE)
max_average_GY <- max(average_GY_by_cultivar$GY)
best_cultivar <- average_GY_by_cultivar$Cultivar[which(average_GY_by_cultivar$GY == max_average_GY)]
cat("Cultivar with the highest average grain yield:", best_cultivar)
```

# d.

Create a table showing the average value of each quantitative variable

```{r}
#| echo: true
vars <- c("PH", "IFP", "NLP", "NGP", "NGL", "NS", "MHG", "GY")
overall_mean <- colMeans(soybean[, vars], na.rm = TRUE)
overall_row <- data.frame(Group = "Overall", t(overall_mean))
season_mean <- aggregate(soybean[, vars], by = list(Season = soybean$Season), FUN = mean, na.rm = TRUE)
season_mean <- data.frame(Group = paste("Season", season_mean$Season), season_mean[, vars])
table_average <- rbind(overall_row, season_mean)
print(table_average)
```

# e.

Carry out a series of t-tests to examine whether the grain yield (GY) differs between the two seasons.

Null hypothesis_H0: The mean grain yields of the two seasons are equal Alternative hypothesis_H1: The mean grain yields of the two seasons are not equal

```{r}
#| echo: true
GY_season1 <- soybean$GY[soybean$Season == 1]
GY_season2 <- soybean$GY[soybean$Season == 2]
t_test_result <- t.test(GY_season1, GY_season2)
t_test_result
```

The p-value is `r round(t_test_result$p.value, 4)`, which is greater than 0.05. Therefore, we fail to reject the null hypothesis. There is no statistically significant difference in grain yield between the two seasons.

## Problem 2 - Adult Income Dataset
# a.
Import the data into a data.frame in R. As with the previous data, you may download the data outside of your submission, but importation should take place inside the problem set submission. ß
```{r}
#| echo: true
adult <- read.table("data/adult/adult.data", sep = ",", header = FALSE)
```

# b.
Clean and simplify the variable names.
```{r}
#| echo: true
names(adult) <- c("age","workclass","fnlwgt","education","education_num","marital_status","occupation","relationship","race","sex","capital_gain","capital_loss","hours_per_week","native_country","income")
```

# c.
Examine missing values. Report the number of missing observations for each variable.
```{r}
#| echo: true
adult[adult == " ?"] <- NA
miss_value <- sapply(adult, function(x) sum(is.na(x)))
print(miss_value)
```

# d.
Restrict the sample to individuals working at least 10 hours per week. Confirm the sample size before and after filtering.
```{r}
#| echo: true
n_before <- nrow(adult)
adult_1 <- adult[adult$hours_per_week >= 10, ]
n_after <- nrow(adult_1)
cat("The sample size before filtering is", n_before,
    "\nThe sample size after filtering is", n_after)
```

# e.
Remove observations with missing values for occupation, workclass, or native country. Explain why.
```{r}
#| echo: true
adult_2 <- adult_1[!is.na(adult_1$occupation) & !is.na(adult_1$workclass) & !is.na(adult_1$native_country),]
```
Use !is.na()to find non-missing values for occupation, workclass and native_country in adult.
Use & to make all three conditions true.
And the requirement only need us to remove observations from these three variates, so we do not to limit columns, so in function "adult[,]", we do not need to write any code after ",".

# f.
Eliminate any observations with: age less than 18, age greater than 90, education level identified as Preschool.
```{r}
#| echo: true
adult_3 <- adult_2[adult_2$age<=90 & adult_2$age>=18 & adult_2$education!="Preschool",]
```

# g.
Report your final sample size.
```{r}
#| echo: true
final_size <- nrow(adult_3)
cat("The final sample size is",final_size)
```

## Problem 3 - Happy Numbers

# a.
Write a function nextHappy() that given a positive integer, computes the next value in the happy-number sequence.
```{r}
#| echo: true
#' nextHappy
#'
#' @param input a positive integer
#'
#' @returns output a positive integer, the sum of the squares of the digits of input
nextHappy <- function(input){
  if(length(input) != 1 | !is.numeric(input) | is.na(input) | input <= 0 || input %% 1 != 0) {
    stop("Input must be a positive integer.")
  }
  output <- 0
  while(input > 0){
    last_digit <- input %% 10
    output <- output + last_digit^2
    input <- input %/% 10
  }
  output
}
nextHappy(19)
nextHappy(82)
```

# b.
Create a function happySequence() that returns the sequence beginning at a given number. Use your nextHappy() function to perform the calculation.
```{r}
#| echo: true
#' Title
#'
#' @param input A positive integer
#'
#' @returns output A list containing: 
#'          sequence: A vector containing the values in the Happy sequence, beginning with the input               value and ending at 1 if the number is happy. 
#'          isHappy: A logical value indicating whether the input is a happy number (TRUE if the                  sequence reaches 1, FALSE otherwise).
happySequence <- function(input){
  if(length(input) != 1 | !is.numeric(input) | is.na(input) | input <= 0 || input %% 1 != 0) {
    stop("Input must be a positive integer.")
  }
  seq <- c(input)
  num <- input
  is_happy <- FALSE
  while (num != 1) {
    num <- nextHappy(num)
    if (num %in% seq) {
      break  
    }
    seq <- c(seq,num)
  }
  if (num == 1) {
    is_happy <- TRUE
  }
  if (is_happy == FALSE) {
    return(list(isHappy = FALSE))
  } else {
    return(list(sequence = seq, isHappy = TRUE))
}}
happySequence(19)
happySequence(4)
```

# c.
Among integers 1 through 1000:
How many are happy?
What percentage are happy?
What starting value produces the longest happy sequence before terminating?
```{r}
#| echo: true
happy <- 0
max_length <- 0
longest_start <- NA

for (i in 1:1000) {
  a <- happySequence(i)
  
  if (a$isHappy == TRUE) {
    happy <- happy + 1
    seq_len <- length(a$sequence)
    if (seq_len > max_length) {
      max_length <- seq_len
      longest_start <- i
    }
  }
}

percent_happy <- happy / 1000 * 100

cat("Number of happy numbers between 1 and 1000 is", happy, "\n")
cat("Percentage of happy numbers is", round(percent_happy, 2), "%\n")
cat("Starting value with the longest happy sequence:", longest_start, "\n")
```

